---
title: "Accessing RW sampler history: proposal scale, covariace, and acceptance rate"
author: Daniel Turek
date: 21 January 2023
output: html_document
---

<!---
danielturek.github.io/public/RWscaleHistory/RWscaleHistory.html
-->

Back to [NIMBLE Vignettes](https://danielturek.github.io/public/nimble_vignettes.html)

\  

NIMBLE's `RW` and `RW_block` samplers contain methods for tracking and accessing the complete history during the course of an MCMC chain of the proposal scale, acceptance rate, and proposal covariance (`RW_block` sampler only).

This vignette demonstrates how to access these these histories.


\  

Specify the two NIMBLE system options:

```{r, message = FALSE}
library(nimble)

nimbleOptions(MCMCsaveHistory = TRUE)
nimbleOptions(buildInterfacesForCompiledNestedNimbleFunctions = TRUE)
```

\   

Build a NIMBLE model and MCMC algorithm.

We'll used the venerated `pump` model, and manually configure the MCMC to also include one `RW_block` sampler.


```{r message = FALSE}
code <- nimbleCode({ 
    alpha ~ dexp(1)
    beta ~ dgamma(0.1, 1)
    for(i in 1:N) {
        theta[i] ~ dgamma(alpha, beta)
        lambda[i] <- theta[i] * t[i]
        x[i] ~ dpois(lambda[i])
    }
})

N <- 10
constants <- list(N = N, t = c(94.3, 15.7, 62.9, 126, 5.24, 31.4, 1.05, 1.05, 2.1, 10.5))
data <- list(x = c(5, 1, 5, 14, 3, 19, 1, 1, 4, 22))
inits <- list(alpha = 1, beta = 1, theta = rep(0.1, N))

Rmodel <- nimbleModel(code, constants, data, inits)

conf <- configureMCMC(Rmodel)

conf$addSampler(target = c('alpha', 'beta'), type = 'RW_block', silent = TRUE)

Rmcmc <- buildMCMC(conf)
```

\   

Now compile the model and the MCMC, and run the MCMC algorithm for a modest
1000 iterations.

```{r message = FALSE}
Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)

set.seed(0)
samples <- runMCMC(Cmcmc, 1000)
```

\   

Now, we'll access the proposal and acceptance rate histories of the
`RW` and `RW_block` samplers.

Use `conf$printSamplers()` to see a printout of all the MCMC samplers.

```{r }
conf$printSamplers()
```

More directly, use the `type = 'RW'` argument to print only `RW`
and `RW_block` samplers.

```{r }
conf$printSamplers(type = 'RW')
```

We see the `RW` sampler on `alpha` is sampler \#1, and the bivariate `RW_block`
operating jointly on `alpha` and `beta` is sampler #13.

The MCMC was run for 1000 iterations, and the default value of the
adapatation interval for NIMBLE's MCMC samplers is every 200
iterations.  Thus, the samplers underwent five adaptations, and the empirical
acceptance rate was recorded five times.

To access the scale and acceptance rate of the `RW` sampler for each
of these adaptive phases:

```{r }
Cmcmc$samplerFunctions$contentsList[[1]]$getScaleHistory()
Cmcmc$samplerFunctions$contentsList[[1]]$getAcceptanceHistory()
```

Similarly, the `RW_block` sampler underwent five adaptations.  In
addition to a scalar `scale` and acceptance rate, the `RW_block`
sampler also has (in this case a 2x2) proposal covariance matrix.  To
access the history of these variables:

```{r }
Cmcmc$samplerFunctions$contentsList[[13]]$getScaleHistory()
Cmcmc$samplerFunctions$contentsList[[13]]$getPropCovHistory()
Cmcmc$samplerFunctions$contentsList[[13]]$getAcceptanceHistory()
```

The proposal covariance history is a 5x2x2 array.  That is, the
initial default proposal covariance matrix was the 2x2 identity
matrix:

```{r }
Cmcmc$samplerFunctions$contentsList[[13]]$getPropCovHistory()[1,,]
```

and the proposal covariance after the final adaptation phase was:

```{r }
Cmcmc$samplerFunctions$contentsList[[13]]$getPropCovHistory()[5,,]
```


\  

We may wish to use these adapated parameters as the starting values
of the sampler tuning parameters in a subsequent (perhaps longer) MCMC run.

Let's extract the final `scale` value of the `RW` sampler, and the
final proposal covariance of the `RW_block` sampler:

```{r }
(RW_scale <- Cmcmc$samplerFunctions$contentsList[[1]]$getScaleHistory()[5])
(RW_block_cov <- Cmcmc$samplerFunctions$contentsList[[13]]$getPropCovHistory()[5,,])
```

\   


Finally, we'll configure and build a new MCMC algorithm, using this
`scale` and proposal covariance as the starting values for the `RW`
and `RW_block` samplers.


```{r }
conf <- configureMCMC(Rmodel)

conf$removeSampler('alpha', print = TRUE)

conf$addSampler(target = 'alpha', type = 'RW', scale = RW_scale, print = TRUE)

conf$addSampler(target = c('alpha', 'beta'), type = 'RW_block',
                propCov = RW_block_cov, silent = TRUE, print = TRUE)

conf$printSamplers()

Rmcmc <- buildMCMC(conf)
```

\   

Voila!




\  

\  

\  

\  

\  

\  

\  

\  

\  

\  

\  

\  

\  

\  

\  

\  

\  






