---
title: Restarting NIMBLE MCMC
author: Daniel Turek
output: html_document
---

<!---
danielturek.github.io/public/restartingMCMC/restartingMCMC.html
-->

\    

This document gives a sketch of how to restart a NIMBLE MCMC algorithm from where it left off, assuming you have to end your R session between runs.  If R didn't have to be restarted, you could simply use the `reset = FALSE` argument to the `Cmcmc$run()` method.  But if the R session must be stopped and restarted, it's a little more complicated.

This approach is not covered in the <a href="https://r-nimble.org/manuals/NimbleUserManual.pdf" target="_blank">NIMBLE User Manual</a>, and requires some knowledge of the internals and implementation of NIMBLE and NIMBLE's MCMC.


\   

## Creating Model and MCMC

```{r, message = FALSE}
library(nimble)
```

Suppose you have the `code`, `constants`, `data`, and `inits`
necessary for creating a NIMBLE `model`, for MCMC sampling.  Here's a
simple example below of Poisson regression.

Note that we'll use exactly these same `code`, `constants`, `data`, and
`inits` after restarting R, to create new `model` and `mcmc` objects.



```{r message = FALSE}
code <- nimbleCode({
    mu ~ dnorm(0, 0.0001)
    sigma ~ dunif(0, 1000)
    for(i in 1:2) {
        beta[i] ~ dnorm(0, 0.0001)
    }
    for(i in 1:N) {
        epsilon[i] ~ dnorm(mu, sd = sigma)
        log(lambda[i]) <- beta[1] + beta[2] * x[i] + epsilon[i]
        y[i] ~ dpois(lambda[i])
    }
})

set.seed(0)
N <- 5
x <- rnorm(N, 0, 2)
lambda <- exp(4 + .3*x)
y <- rpois(N, lambda)

constants <- list(N = N, x = x)
data <- list(y = y)
inits <- list(mu = 0, sigma = 1, beta = rep(0,2), epsilon = rep(0, N))

Rmodel <- nimbleModel(code, constants, data, inits)
```

For good measure, run `Rmodel$calculate()` to make sure the model is
fully initialized.

```{r }
Rmodel$calculate()
```



Now we'll create and configure the MCMC algorithm.  To make for a more
interesting example,  demonstrating how to handle different samplers,
we'll assign a `slice`
sampler on `sigma`.

We'll leave the
conjugate sampler that is assigned to `mu`, and the`RW` samplers
assigned to the `beta` and `epsilon` terms.

```{r }
conf <- configureMCMC(Rmodel)
conf$printSamplers()
conf$removeSamplers("sigma")
conf$addSampler("sigma", "slice")
conf$printSamplers()
```

That looks good.  We'll use this MCMC.

Next, we'll build the MCMC, compile the model and the MCMC, and run
the MCMC for 10,000 iterations.

```{r message = FALSE}
Rmcmc <- buildMCMC(conf)

Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)

set.seed(0)
Cmcmc$run(10000)
samples <- as.matrix(Cmcmc$mvSamples)
```

\   

## What to Save

Now, before ending the R session, we'll need to save several things:

1. The MCMC samples we've already run.
2. The values of all non-data stochastic nodes in the model.  That is,
  anything in the model `code` defined using `~`, which is not
  specified as `data`.  In this case, that's `mu`, `sigma`, `beta`, and
  `epsilon`.
3. The internal state variables from all the MCMC samplers.  This will
  be the slightly tricky part, but not too difficult.

\   

##### Samples

Straight forward, make sure to save the MCMC `samples` that you've
already run.

\   

##### Model Variables

For any parameters or latent variables that are being sampled in the
model, you'll need to save the "current" values of these which are
stored in the `Cmodel` object.  These can be accessed from the
`Cmodel` object using the syntax `Cmodel$variableName`.

I'll assume you know for your model what these are, but you can also
look at everything that's being sampled
using. `conf$printSamplers()`.  Another way to tell would be:

```{r }
Cmodel$getNodeNames(stochOnly = TRUE, includeData = FALSE)
```

Or even nicer:

```{r }
nodes <- Cmodel$getNodeNames(stochOnly = TRUE, includeData = FALSE)
Cmodel$getVarNames(nodes = nodes)
```

So, we'll extract these and store them.  Let's use a
list to hold the model variables.

```{r }
modelVariables <- list()

modelVariables$mu <- Cmodel$mu
modelVariables$sigma <- Cmodel$sigma
modelVariables$beta <- Cmodel$beta
modelVariables$epsilon <- Cmodel$epsilon
```

This list should now contain the values of all non-data stochastic
variables in the model:

```{r }
modelVariables
```

You'll want to save this list `modelVariables` object.

\   

##### MCMC Sampler State Variables

NIMBLE's MCMC samplers usually contain one or more "state variables",
which define the state of the sampler algorithm, and are updated as
the MCMC progresses.  You'll need to extract and save the values of
these state variables for each sampling algorithm.

We'll use an R `list` to hold the state variables.  The first element
will itself be a list of the state variables for the first sampler.
The second element will record the state variables from the second
sampler, etc.

```{r }
samplerStates <- list()
```

Again, you can see how many samplers there are, and what types they
are, using:


```{r }
conf$printSamplers()
```

The first sampler is a `conjugate` sampler (assigned to `mu`). 
Conjugate samplers don't require any internal state.   So the first 
element in our list (the state variables for the first sampler) won't 
require anything. 

To tell what the "state variables" for any other sampler are,
your best bet is looking at the
underlying code for each sampler, available in <a
href="https://github.com/nimble-dev/nimble/blob/devel/packages/nimble/R/MCMC_samplers.R"
target="_blank">MCMC_samplers.R</a>.

For each sampling algorithm, for example the `RW` sampler, find
corresponding `nimbleFunction` definition, in this case `sampler_RW`.
Near the end of the definition there will be a `reset` method.  In
this case, for the `RW` sampler, this looks like:

```
reset = function() {
    scale <<- scaleOriginal 
    timesRan      <<- 0
    timesAccepted <<- 0
    timesAdapted  <<- 0
    ##scaleHistory  <<- scaleHistory * 0    ## scaleHistory
    gamma1 <<- 0
}
```

The "state variables" for the sampler are all the variables that are
reset to some initial value in this function.  So in this case:
`scale`, `timesRan`, `timesAccepted`, `timesAdapted`, and `gamma1`.  For each of these, you'll need to extract the value of this variable
from the compiled sampler algorithm, and save it.

There happen to be
seven `RW` samplers (two for the `beta` terms, and one for each
`epsilon`) so we'll use a loop to extract these state variables from
each compiled sampler algorithm, and store them into the appropriate
element of our list.  Loop over sampler numbers 2 through 8.

```{r }
for(i in 2:8) {
    samplerStates[[i]] <- list()
    samplerStates[[i]]$scale <- valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "scale")
    samplerStates[[i]]$timesRan <- valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "timesRan")
    samplerStates[[i]]$timesAccepted <- valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "timesAccepted")
    samplerStates[[i]]$timesAdapted <- valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "timesAdapted")
    samplerStates[[i]]$gamma1 <- valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "gamma1")
}
```

For the 9th sampler, a `slice`sampler on `sigma`, the
code for the `sampler_slice` `reset()` method looks like:


```
reset = function() {
    width        <<- widthOriginal
    timesRan     <<- 0
    timesAdapted <<- 0
    sumJumps     <<- 0
}
```

So, we'll store these four state variables:


```{r }
samplerStates[[9]] <- list()
samplerStates[[9]]$width <- valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[9]], "width")
samplerStates[[9]]$timesRan <- valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[9]], "timesRan")
samplerStates[[9]]$timesAdapted <- valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[9]], "timesAdapted")
samplerStates[[9]]$sumJumps <- valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[9]], "sumJumps")
```

Now, the `samplerStates` list will also need to be saved.

\   

## New R Session: Restoring your MCMC

Easy from here.  Say you've now restarted R, and restored the
`samples`, `modelVariables`, and `samplerStates` objects.

First, recreate your model and MCMC.  This code is exactly the same as
above:


```{r message = FALSE}
library(nimble)

code <- nimbleCode({
    mu ~ dnorm(0, 0.0001)
    sigma ~ dunif(0, 1000)
    for(i in 1:2) {
        beta[i] ~ dnorm(0, 0.0001)
    }
    for(i in 1:N) {
        epsilon[i] ~ dnorm(mu, sd = sigma)
        log(lambda[i]) <- beta[1] + beta[2] * x[i] + epsilon[i]
        y[i] ~ dpois(lambda[i])
    }
})

set.seed(0)
N <- 5
x <- rnorm(N, 0, 2)
lambda <- exp(4 + .3*x)
y <- rpois(N, lambda)

constants <- list(N = N, x = x)
data <- list(y = y)
inits <- list(mu = 0, sigma = 1, beta = rep(0,2), epsilon = rep(0, N))

Rmodel <- nimbleModel(code, constants, data, inits)

Rmodel$calculate()

conf <- configureMCMC(Rmodel)
conf$removeSamplers("sigma")
conf$addSampler("sigma", "slice")
conf$printSamplers()

Rmcmc <- buildMCMC(conf)

Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)
```

\   

##### Restore the model variables

Restore the variable values into the `Cmodel` object, from our saved
`modelVariables` list.

The syntax is just the
reverse of extracting them earlier.


```{r }
Cmodel$mu <- modelVariables$mu
Cmodel$sigma <- modelVariables$sigma
Cmodel$beta <- modelVariables$beta
Cmodel$epsilon <- modelVariables$epsilon
```

\    


##### Restore the sampler state variables

Now extract the values from the `samplerStates` list, and put them into the
`Cmcmc` sampler functions.

Here, the syntax is slighly different:


```{r eval = FALSE}
for(i in 2:8) {
    valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "scale", samplerStates[[i]]$scale)
    valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "timesRan", samplerStates[[i]]$timesRan)
    valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "timesAccepted", samplerStates[[i]]$timesAccepted)
    valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "timesAdapted", samplerStates[[i]]$timesAdapted)
    valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "gamma1", samplerStates[[i]]$gamma1)
}

valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[9]], "width", samplerStates[[9]]$width)
valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[9]], "timesRan", samplerStates[[9]]$timesRan)
valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[9]], "timesAdapted", samplerStates[[9]]$timesAdapted)
valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[9]], "sumJumps", samplerStates[[9]]$sumJumps)
```

This has now restored the "state" into the MCMC sampler
algorithms.

\   

## Continue Running MCMC

You can now continue running the `Cmcmc` algorithm, and it will
effectively "pick up where it left off".

Just make sure to use the
`reset = FALSE` argument, so it doesn't reset all the samplers back to
their initial states.


```{r }
set.seed(0)
Cmcmc$run(10000, reset = FALSE)
samples_continued <- as.matrix(Cmcmc$mvSamples)
```

\   

Let me know if anything here is unclear.

Good luck!

Daniel


\   

\   

\   

\   

\   

\   

\   

\   

\   

\   

\   

\   













