
---
title: Avoid recompiling models and MCMCs
author: Daniel Turek
date: 20 September 2016
output: html_document
---

<!---
danielturek.github.io/public/folder/fileName.Rmd
-->

Back to [NIMBLE Vignettes](https://danielturek.github.io/public/nimble_vignettes.html)

\

This demonstrates the procedure to avoid recompiling models and MCMC
algorithms, when experimenting with multiple MCMC algorithms for the
same model.  It attempts to be  general in terms of models and
use cases.

There is, however, a lack of generality for the case when the model contains no
continuous-valued nodes.  We assume that there's at least one
continuous-valued node in the model.

```{r, message = FALSE}
library(nimble)
```

Assume a user has created a model already.  In fact, they've also already created an MCMC for it, and compiled both the model and the MCMC algorithm.


```{r, message = FALSE}
## contains both discrete and continuous nodes
code <- nimbleCode({
    a ~ dbern(0.5)
    b ~ dnorm(0, 1)
    c ~ dnorm(b, 1)
})

Rmodel_user <- nimbleModel(code, inits = list(a=0, b=0, c=0))

conf_user <- configureMCMC(Rmodel_user)
conf_user$printSamplers()

Rmcmc_user <- buildMCMC(conf_user)

Cmodel_user <- compileNimble(Rmodel_user)

Cmcmc_user <- compileNimble(Rmcmc_user, project = Rmodel_user)
```

Now, the user wants to use some experiemental MCMC algorithm on their
model.  They will provide their model object (either `Rmodel_user` or
`Cmodel_user`) as an input to the algorithm.  We don't know which
they'll provide, but we'll handle both cases.

Let's say they choose to provide the compiled model as input,
`Cmodel_user`.


```{r }
model_input <- Cmodel_user
```

The following are all the steps to create and use multiple MCMC
algorithms.

\ 

# One-time steps

\ 

### Recover the Rmodel object

If they provided the compiled (C) model object, then we recover the
uncompiled (R) object:


```{r }
if(inherits(model_input, 'CmodelBaseClass')) 
    model_input <- model_input$Rmodel

## model_input is now an uncompiled model object
inherits(model_input, 'RmodelBaseClass')
```

\ 

### Create a new copy of the model

Replicate the model.  This will be for use in the experimental algorithm.


```{r, message = FALSE}
Rmodel <- model_input$newModel(replicate = TRUE, check = FALSE) 
```

\ 

### Create an initial MCMC configuration

Create an initial configuration object.  We'll need this later.

Also add monitors for all stochastic nodes.


```{r }
conf_initial <- configureMCMC(Rmodel)

monitorsVector <- Rmodel$getNodeNames(stochOnly = TRUE, includeData = FALSE)
conf_initial$addMonitors(monitorsVector, print = FALSE)

conf_initial$getMonitors()
```

\ 

### Add samplers to the initial MCMC configuration

This is important.  You need to add one of each sampler that you will
wish to use, anytime later during the course of the experimental
algorithm, to the initial configuartion.  So you must
know all the different types of samplers you might wish to use.  One
of each will be added to the initial configuration object.

You'll have to be careful here about adding samplers to the
appropriate type of node (e.g., continuous, discrete).  That's why
things can fall apart a little bit if there are no continuous nodes,
since `RW` and `RW_block` samplers can only operate on continuous nodes.

For the purpose of adding those types of samplers, we'll make sure to
extract the name of a **continuous**-valued node.


```{r }
scalarNodeVector <- Rmodel$getNodeNames(stochOnly=TRUE, includeData=FALSE, returnScalarComponents=TRUE)
discreteInd <- sapply(scalarNodeVector, function(n) Rmodel$isDiscrete(n), USE.NAMES=FALSE)
scalarNodeVectorContinuous <<- scalarNodeVector[!discreteInd]
scalarNodeVectorContinuous
```

We'll add the RW and RW_block samplers to the first scalar-valued
node.

This part will break if there are no continuous-valued nodes.

```{r }
firstScalarNode <- scalarNodeVectorContinuous[1]
firstScalarNode
```

\ 

### Add one of each sampler we might use (anytime later)

You'll need to have a list of all the types of samplers that you might
want to use, anytime later in the experimental algorithm.

We'll add one of each to the initial configuration object.

Also, leave in all the samplers already in the initial configuration,
since that will include everything in the default configuration
object.  Most importantly, this will include any conjugate samplers.

```{r }
conf_initial$printSamplers() 

samplersWeMightUse <- c('RW', 'slice', 'RW_block') 

for(sampler in samplersWeMightUse) 
    conf_initial$addSampler(target = firstScalarNode, type = sampler) 

conf_initial$printSamplers() 
```

\ 

### Build and compile the model and initial MCMC

Now we have to compile both our version of the model, and the initial
MCMC algorithm.

These compilations must take place, but only once.  This is a "one-time" step at the beginning of the experimental algorithm.


```{r, message = FALSE}
Rmcmc_initial <- buildMCMC(conf_initial) 

Cmodel <- compileNimble(Rmodel) 
Cmcmc_initial <- compileNimble(Rmcmc_initial, project = Rmodel) 
```

\ 

# Steps to repeat for each new MCMC

Now, we can create and run new MCMC algorithms for the same model,
without a lengthy compilation.  There will be a call to
`compileNimble()`, but it will be fast, and reuse existing C++ code
already created, so it only has to instantiate and populate C++
objects.  It's very quick.

We'll also be careful to start our version of the (compiled) model at
the same initial values every time, just so every MCMC algorithm
begins at the same initial values.

\ 

### Step 1: Set values in Cmodel back to initial values

```{r }
nimCopy(from = model_input, to = Cmodel, logProb = TRUE) 
calculate(Cmodel) 
```

\ 

### Step 2: Create a new configuration object, add any samplers

This step is critical.  Notice we'll use the argument
`configureMCMC(oldConf = ...)`.  That's the key.


```{r }
## create a new configuration from the initial confuration (important!)
conf_new <- configureMCMC(oldConf = conf_initial) 

## remove all samplers
conf_new$setSamplers()
conf_new$printSamplers()
```


Now add whatever samplers out experimental algorithms dictates.

These must be a subset of the types of samplers already compiled.

Say we want all `slice` samplers the first time.

```{r }
nodes <- c('a', 'b', 'c') 

for(node in nodes) 
    conf_new$addSampler(target = node, type = 'slice') 

conf_new$printSamplers() 
```

\ 

### Step 3: Build, compile and run the new MCMC

```{r, message = FALSE}
Rmcmc_new <- buildMCMC(conf_new) 

Cmcmc_new <- compileNimble(Rmcmc_new, project = Rmodel) 

Cmcmc_new$run(1000, progressBar = FALSE)

samples <- as.matrix(Cmcmc_new$mvSamples)
```

\ 

# One more time

Let's do one more "iteration" of our experimental algorithm.

This time, all the repeated steps will be together, and we'll add one
`RW_block` sampler instead of three `slice` samplers.


```{r }
## reset values in Cmodel to initial values
nimCopy(from = model_input, to = Cmodel, logProb = TRUE) 
calculate(Cmodel) 

## create a new configuration from the initial configuration (important!)
conf_new <- configureMCMC(oldConf = conf_initial)

## remove all samplers
conf_new$setSamplers()
conf_new$printSamplers() 

## this time we only want a RW_block sampler on 'b' and 'c'
conf_new$addSampler(target = c('b', 'c'), type = 'RW_block') 

conf_new$printSamplers() 

## build, compile, and run
Rmcmc_new <- buildMCMC(conf_new) 
Cmcmc_new <- compileNimble(Rmcmc_new, project = Rmodel) 
Cmcmc_new$run(1000, progressBar = FALSE)

## extract samples for analysis
samples <- as.matrix(Cmcmc_new$mvSamples)
```

\ 

Make sense?

Questions or comments to <dbt1@williams.edu>

\

\

\










