
---
title: Adding a Conjugate Sampler
author: Daniel Turek
output: html_document
---


<!---
danielturek.github.io/public/folder/fileName.Rmd
-->


It turns out there's already a function (already on devel, Chris!)
that handles this for you.

If you can believe it, it's called `addConjugateSampler()`.

```{r, message = FALSE}
library(nimble)
```

Create a model with a conjugate node.

And a non-conjugate node, for the sake of example.


```{r }

code <- nimbleCode({
     theta ~ dbeta(1, 1)   ## theta will be conjugate
     for(i in 1:2)
         x[i] ~ dbin(prob = theta, size = 10)
     for(i in 1:3)
         y[i] ~ dnegbin(prob = theta, size = 50)
     notConjugate ~ dbeta(1, 1)    ## notConjugate will not be conjugate
     z ~ dnorm(notConjugate, 1)
})

Rmodel <- nimbleModel(code)
```


Say you want to programmatically add a conjugate sampler for `theta`,
which is conjugate.

```{r }
node <- 'theta'

conjInfo <- Rmodel$checkConjugacy2(node)[[node]]

## the conjInfo object has everything we need:
conjInfo

conf <- configureMCMC(Rmodel, nodes = NULL)

conf$printSamplers()  ## initially, no samplers

conf$addConjugateSampler(conjInfo)

conf$printSamplers()  ## now we have the correct conjugate sampler for 'theta'

```

If you try this procedure with a node that's *not* conjugate:


```{r }

node <- 'notConjugate'   ## this node is not conjugate

conjInfo <- Rmodel$checkConjugacy2(node)[[node]]

## conjInfo should be NULL
conjInfo

```

Let me know if you have any questions.

\

\



