---
title: Using MCMC Samples as a Prior Distribution
author: Daniel Turek
date: 1 October 2023
output: html_document
---

<!---
danielturek.github.io/public/prior_samples/prior_samples.html
-->

Back to [NIMBLE Vignettes](https://danielturek.github.io/public/nimble_vignettes.html)

\    

Here, we introduce how to use a set of MCMC samples (for a model parameter) to define the empirical prior distribution of a parameter in a subsequent model and/or MCMC.  This is accomplised using the `prior_samples` MCMC sampler, defined below.  A more refined version of this sampler may later be introduced into the `nimble` package itself.

\   

## Define the `prior_samples` MCMC sampler

```{r, message = FALSE}
library(nimble)

sampler_prior_samples <- nimbleFunction(
    name = 'sampler_prior_samples',
    contains = sampler_BASE,
    setup = function(model, mvSaved, target, control) {
        ## control list extraction
        samples     <- extractControlElement(control, 'samples',     error = 'prior_samples sampler missing required control argument: samples')
        randomDraws <- extractControlElement(control, 'randomDraws', FALSE)
        ## node list generation
        targetExpanded <- model$expandNodeNames(target)
        targetAsScalar <- model$expandNodeNames(targetExpanded, returnScalarComponents = TRUE)
        ccList <- nimble:::mcmc_determineCalcAndCopyNodes(model, targetExpanded)
        calcNodes <- ccList$calcNodes; calcNodesNoSelf <- ccList$calcNodesNoSelf; copyNodesDeterm <- ccList$copyNodesDeterm; copyNodesStoch <- ccList$copyNodesStoch
        ## numeric value generation
        k <- length(targetAsScalar)
        if(is.null(dim(samples)))   samples <- matrix(samples, ncol = 1)    ## make vectors into 1-column array
        nSamples <- dim(samples)[1]
        ind <- 0
        ## checks
        if(length(dim(samples)) != 2)   stop(paste0('  [Error] prior_samples sampler \'samples\' control argument must be a 2-dimensional array, but value provided was a ', length(dim(samples)), '-dimensional array'), call. = FALSE)
        if(!(storage.mode(samples) %in% c('integer', 'double')))   stop('  [Error] prior_samples sampler \'samples\' control argument must be numeric or integer type', call. = FALSE)
        if(dim(samples)[2] != k)   stop(paste0('  [Error] prior_samples sampler \'samples\' control argument had ', dim(samples)[2], ' columns, but target nodes have ', k, ' scalar elements.  These numbers must be equal.'), call. = FALSE)
    },
    run = function() {
        if(randomDraws) {
            ind <<- ceiling(runif(1, 0, nSamples))   ## random draws
        } else {
            ind <<- ind + 1                          ## sequential draws (the default)
            if(ind > nSamples)   ind <<- 1           ## recycle sequential draws, if necessary
        }
        ##model$values(targetExpanded) <<- samples[ind, 1:k]   ## XXX?????
        values(model, targetExpanded) <<- samples[ind, 1:k]
        model$calculate(calcNodes)
        nimCopy(from = model, to = mvSaved, row = 1, nodes = target, logProb = FALSE)
        nimCopy(from = model, to = mvSaved, row = 1, nodes = copyNodesDeterm, logProb = FALSE)
        nimCopy(from = model, to = mvSaved, row = 1, nodes = copyNodesStoch, logProbOnly = TRUE)
    },
    methods = list(
        before_chain = function(MCMCniter = double(), MCMCnburnin = double(), MCMCchain = double()) {
            ## issue a note if sequential draws from prior samples will have to be recycled
            if((!randomDraws) & (MCMCniter > nSamples))   print('  [Note] prior_samples sampler will recycle sequential draws from prior samples, since ', nSamples, ' samples were provided, and ', MCMCniter, ' MCMC iterations will be run.')
        },
        reset = function() {
            ind <<- 0
        }
    )
)

code <- nimbleCode({
    a ~ dnorm(0, 1)
    b ~ dnorm(0, 1)
    for(i in 1:10) {
        rr[i] <- r[i]^2
        x[i] ~ dnorm(rr[i] + a, s[i] + b)
    }
})

inits <- list(a=1, b=1, x=rep(0,10))

Rmodel <- nimbleModel(code, inits = inits)

conf <- configureMCMC(Rmodel, nodes = NULL, monitors = c('a', 'b', 'r', 'rr', 's'))
##conf$addSampler('a', 'prior_samples', samples = 1:10)
##conf$addSampler('b', 'prior_samples', samples = (1:10)/10)
conf$addSampler(c('a', 'b', 'r[1]'), 'prior_samples', samples = cbind(1:10, (1:10)/10, 2))
conf$printSamplers()

Rmcmc <- buildMCMC(conf)

set.seed(0)
samples <- runMCMC(Rmcmc, 10)

samples

compiledList <- compileNimble(list(model=Rmodel, mcmc=Rmcmc))
Cmodel <- compiledList$model; Cmcmc <- compiledList$mcmc
##Cmodel <- compileNimble(Rmodel)
##Cmcmc <- compileNimble(Rmcmc, project = Rmodel)#, showCompilerOutput = TRUE)

set.seed(0)
samples <- runMCMC(Cmcmc, 10000)





```

Suppose you have the `code`, `constants`, `data`, and `inits`
necessary for creating a NIMBLE `model`, for MCMC sampling.  Here's a
simple example below of Poisson regression.

Note that we'll use exactly these same `code`, `constants`, `data`, and
`inits` after restarting R, to create new `model` and `mcmc` objects.



```{r message = FALSE}
code <- nimbleCode({
    mu ~ dnorm(0, 0.0001)
    sigma ~ dunif(0, 1000)
    for(i in 1:2) {
        beta[i] ~ dnorm(0, 0.0001)
    }
    for(i in 1:N) {
        epsilon[i] ~ dnorm(mu, sd = sigma)
        log(lambda[i]) <- beta[1] + beta[2] * x[i] + epsilon[i]
        y[i] ~ dpois(lambda[i])
    }
})

set.seed(0)
N <- 5
x <- rnorm(N, 0, 2)
lambda <- exp(4 + .3*x)
y <- rpois(N, lambda)

constants <- list(N = N, x = x)
data <- list(y = y)
inits <- list(mu = 0, sigma = 1, beta = rep(0,2), epsilon = rep(0, N))

Rmodel <- nimbleModel(code, constants, data, inits)
```

For good measure, run `Rmodel$calculate()` to make sure the model is
fully initialized.

```{r }
Rmodel$calculate()
```



Now we'll create and configure the MCMC algorithm.  To make for a more
interesting example,  demonstrating how to handle different samplers,
we'll assign a `slice`
sampler on `sigma`.

We'll leave the
conjugate sampler that is assigned to `mu`, and the`RW` samplers
assigned to the `beta` and `epsilon` terms.

You can decide whether or not to include WAIC by setting `useWAIC`.

```{r }
useWAIC <- TRUE
conf <- configureMCMC(Rmodel, enableWAIC = useWAIC)
conf$printSamplers()
conf$removeSamplers("sigma")
conf$addSampler("sigma", "slice")
conf$printSamplers()
```

That looks good.  We'll use this MCMC.

Next, we'll build the MCMC, compile the model and the MCMC, and run
the MCMC for 10,000 iterations.

```{r message = FALSE}
Rmcmc <- buildMCMC(conf)

Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)

set.seed(0)
Cmcmc$run(10000)
samples <- as.matrix(Cmcmc$mvSamples)
```

\   

## What to Save

Now, before ending the R session, we'll need to save several things:

1. The MCMC samples we've already run.
2. The values of all non-data stochastic nodes in the model.  That is,
  anything in the model `code` defined using `~`, which is not
  specified as `data`.  In this case, that's `mu`, `sigma`, `beta`, and
  `epsilon`.
3. The internal state variables from all the MCMC samplers.  This will
be the slightly tricky part, but not too difficult.
4. R's random number seed.
5. If using WAIC, the internal state variables for the WAIC calculation.

\   

##### Samples

Straight forward, make sure to save the MCMC `samples` that you've
already run.

\   

##### Model Variables

For any parameters or latent variables that are being sampled in the
model, you'll need to save the "current" values of these which are
stored in the `Cmodel` object.  These can be accessed from the
`Cmodel` object using the syntax `Cmodel$variableName`.

I'll assume you know for your model what these are, but you can also
look at everything that's being sampled
using. `conf$printSamplers()`.  Another way to tell would be:

```{r }
Cmodel$getNodeNames(stochOnly = TRUE, includeData = FALSE)
```

Or even nicer:

```{r }
nodes <- Cmodel$getNodeNames(stochOnly = TRUE, includeData = FALSE)
Cmodel$getVarNames(nodes = nodes)
```

So, we'll extract these and store them.  Let's use a
list to hold the model variables.

```{r }
modelVariables <- list()

modelVariables$mu <- Cmodel$mu
modelVariables$sigma <- Cmodel$sigma
modelVariables$beta <- Cmodel$beta
modelVariables$epsilon <- Cmodel$epsilon
```

This list should now contain the values of all non-data stochastic
variables in the model:

```{r }
modelVariables
```

You'll want to save this list `modelVariables` object.

\   

##### MCMC Sampler State Variables  
 
NIMBLE's MCMC samplers usually contain one or more "state variables",
which define the state of the sampler algorithm, and are updated as
the MCMC progresses.  You'll need to extract and save the values of
these state variables for each sampling algorithm.

We'll use an R `list` to hold the state variables.  The first element
will itself be a list of the state variables for the first sampler.
The second element will record the state variables from the second
sampler, etc.

```{r }
samplerStates <- list()
```

Again, you can see how many samplers there are, and what types they
are, using:


```{r }
conf$printSamplers()
```

The first sampler is a `conjugate` sampler (assigned to `mu`). 
Conjugate samplers don't require any internal state.   So the first 
element in our list (the state variables for the first sampler) won't 
require anything. 

To tell what the "state variables" for any other sampler are,
your best bet is looking at the
underlying code for each sampler, available in <a
href="https://github.com/nimble-dev/nimble/blob/devel/packages/nimble/R/MCMC_samplers.R"
target="_blank">MCMC_samplers.R</a>.

For each sampling algorithm, for example the `RW` sampler, find
corresponding `nimbleFunction` definition, in this case `sampler_RW`.
Near the end of the definition there will be a `reset` method.  In
this case, for the `RW` sampler, this looks like:

```
reset = function() {
    scale <<- scaleOriginal 
    timesRan      <<- 0
    timesAccepted <<- 0
    timesAdapted  <<- 0
    ##scaleHistory  <<- scaleHistory * 0    ## scaleHistory
    gamma1 <<- 0
}
```

The "state variables" for the sampler are all the variables that are
reset to some initial value in this function.  So in this case:
`scale`, `timesRan`, `timesAccepted`, `timesAdapted`, and `gamma1`.  For each of these, you'll need to extract the value of this variable
from the compiled sampler algorithm, and save it.

There happen to be
seven `RW` samplers (two for the `beta` terms, and one for each
`epsilon`) so we'll use a loop to extract these state variables from
each compiled sampler algorithm, and store them into the appropriate
element of our list.  Loop over sampler numbers 2 through 8.

```{r }
for(i in 2:8) {
    samplerStates[[i]] <- list()
    samplerStates[[i]]$scale <- valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "scale")
    samplerStates[[i]]$timesRan <- valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "timesRan")
    samplerStates[[i]]$timesAccepted <- valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "timesAccepted")
    samplerStates[[i]]$timesAdapted <- valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "timesAdapted")
    samplerStates[[i]]$gamma1 <- valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "gamma1")
}
```

For the 9th sampler, a `slice`sampler on `sigma`, the
code for the `sampler_slice` `reset()` method looks like:


```
reset = function() {
    width        <<- widthOriginal
    timesRan     <<- 0
    timesAdapted <<- 0
    sumJumps     <<- 0
}
```

So, we'll store these four state variables:


```{r }
samplerStates[[9]] <- list()
samplerStates[[9]]$width <- valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[9]], "width")
samplerStates[[9]]$timesRan <- valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[9]], "timesRan")
samplerStates[[9]]$timesAdapted <- valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[9]], "timesAdapted")
samplerStates[[9]]$sumJumps <- valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[9]], "sumJumps")
```

Now, the `samplerStates` list will also need to be saved.

\   


##### R's random number seed

We'll also save the current value of R's random number generation
seed, so we can pick up the MCMC sampling precisely as it would have
continued, had we actually kept running the MCMC chain now.


```{r }
seed <- .Random.seed
```

##### WAIC State Variables  

If you're using WAIC, you'll also need to save the current values of some WAIC-related quantities that are used in NIMBLE's online WAIC algorithm.

```{r }
waicState <- list()
if(useWAIC) {
    waicState$mcmcIter <- valueInCompiledNimbleFunction(Cmcmc$waicFun[[1]], "mcmcIter")
    waicState$lppdSumMaxMat <- valueInCompiledNimbleFunction(Cmcmc$waicFun[[1]], "lppdSumMaxMat")
    waicState$lppdCurSumMat <- valueInCompiledNimbleFunction(Cmcmc$waicFun[[1]], "lppdCurSumMat")
    waicState$sspWAICmat <- valueInCompiledNimbleFunction(Cmcmc$waicFun[[1]], "sspWAICmat")
    waicState$meanpWAICmat <- valueInCompiledNimbleFunction(Cmcmc$waicFun[[1]], "meanpWAICmat")
    waicState$delta1pWAICmat <- valueInCompiledNimbleFunction(Cmcmc$waicFun[[1]], "delta1pWAICmat")
    waicState$delta2pWAICmat <- valueInCompiledNimbleFunction(Cmcmc$waicFun[[1]], "delta2pWAICmat")
    waicState$logProbMat <- valueInCompiledNimbleFunction(Cmcmc$waicFun[[1]], "logProbMat")
    waicState$finalized <- valueInCompiledNimbleFunction(Cmcmc$waicFun[[1]], "finalized")
}
```

\   

Now, save all these to a file.


```{r }
fileName <- tempfile()
save(samples, modelVariables, samplerStates, waicState, seed, file = fileName)
```



\   


## New R Session: Restoring your MCMC  
 
Easy from here.  Say you've now restarted R, and restored the
`samples`, `modelVariables`, and `samplerStates` objects.

First, recreate your model and MCMC.  This code is exactly the same as
above:


```{r message = FALSE}
###########################
### Start new R session ###
###########################

library(nimble)

code <- nimbleCode({
    mu ~ dnorm(0, 0.0001)
    sigma ~ dunif(0, 1000)
    for(i in 1:2) {
        beta[i] ~ dnorm(0, 0.0001)
    }
    for(i in 1:N) {
        epsilon[i] ~ dnorm(mu, sd = sigma)
        log(lambda[i]) <- beta[1] + beta[2] * x[i] + epsilon[i]
        y[i] ~ dpois(lambda[i])
    }
})

set.seed(0)
N <- 5
x <- rnorm(N, 0, 2)
lambda <- exp(4 + .3*x)
y <- rpois(N, lambda)

constants <- list(N = N, x = x)
data <- list(y = y)
inits <- list(mu = 0, sigma = 1, beta = rep(0,2), epsilon = rep(0, N))

Rmodel <- nimbleModel(code, constants, data, inits)

Rmodel$calculate()

conf <- configureMCMC(Rmodel, enableWAIC = useWAIC)
conf$removeSamplers("sigma")
conf$addSampler("sigma", "slice")
conf$printSamplers()

Rmcmc <- buildMCMC(conf)

Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)
```

Then load our saved file:

```{r }
##fileName <- '~/temp/restartMCMC.RData'
load(fileName)
```

\   

##### Restore the model variables

Restore the variable values into the `Cmodel` object, from our saved
`modelVariables` list.





The syntax is just the
reverse of extracting them earlier.


```{r }
Cmodel$mu <- modelVariables$mu
Cmodel$sigma <- modelVariables$sigma
Cmodel$beta <- modelVariables$beta
Cmodel$epsilon <- modelVariables$epsilon
```

The following code block is new. A few things were overlooked before, since the
`reset = FALSE` option was designed to continue running an MCMC that had
already been run.   The specific issues are pretty technical,
but have to do with the `mvSaved` and `mvSamples` data
structures internal to the `Cmcmc` algorithm, and initializing those
datastructures correctly.

These lines take care of it:

```{r }
Cmodel$calculate()

for(name in names(Cmcmc$mvSaved$sizes)) {
    Cmcmc$mvSaved[[name]] <- Cmodel[[name]]
}

resize(Cmcmc$mvSamples, 0)
```

\    


##### Restore the sampler state variables  
 
Now extract the values from the `samplerStates` list, and put them into the
`Cmcmc` sampler functions.

Here, the syntax is slighly different:


```{r eval = FALSE}
for(i in 2:8) {
    valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "scale", samplerStates[[i]]$scale)
    valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "timesRan", samplerStates[[i]]$timesRan)
    valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "timesAccepted", samplerStates[[i]]$timesAccepted)
    valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "timesAdapted", samplerStates[[i]]$timesAdapted)
    valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[i]], "gamma1", samplerStates[[i]]$gamma1)
}

valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[9]], "width", samplerStates[[9]]$width)
valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[9]], "timesRan", samplerStates[[9]]$timesRan)
valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[9]], "timesAdapted", samplerStates[[9]]$timesAdapted)
valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[9]], "sumJumps", samplerStates[[9]]$sumJumps)
```

This has now restored the "state" into the MCMC sampler
algorithms.

\   


##### Restore the sampler state variables 

And, we'll restore R's random number generation seed to where it left
off.


```{r }
.Random.seed <- seed
```

\   

##### Restore WAIC state variables 

If using WAIC, restore the WAIC state variables.

```{r}
if(useWAIC) {
    valueInCompiledNimbleFunction(Cmcmc$waicFun[[1]], "mcmcIter", waicState[["mcmcIter"]])
    valueInCompiledNimbleFunction(Cmcmc$waicFun[[1]], "lppdSumMaxMat", waicState[["lppdSumMaxMat"]])
    valueInCompiledNimbleFunction(Cmcmc$waicFun[[1]], "lppdCurSumMat", waicState[["lppdCurSumMat"]])
    valueInCompiledNimbleFunction(Cmcmc$waicFun[[1]], "sspWAICmat", waicState[["sspWAICmat"]])
    valueInCompiledNimbleFunction(Cmcmc$waicFun[[1]], "meanpWAICmat", waicState[["meanpWAICmat"]])
    valueInCompiledNimbleFunction(Cmcmc$waicFun[[1]], "delta1pWAICmat", waicState[["delta1pWAICmat"]])
    valueInCompiledNimbleFunction(Cmcmc$waicFun[[1]], "delta2pWAICmat", waicState[["delta2pWAICmat"]])
    valueInCompiledNimbleFunction(Cmcmc$waicFun[[1]], "logProbMat", waicState[["logProbMat"]])
    valueInCompiledNimbleFunction(Cmcmc$waicFun[[1]], "finalized", waicState[["finalized"]])
}                                                 
```

\

## Continue Running MCMC

You can now continue running the `Cmcmc` algorithm, and it will
effectively "pick up where it left off".

Just make sure to use the
`reset = FALSE` argument, so it doesn't reset all the samplers back to
their initial states.

And if using WAIC, similarly set `resetWAIC = FALSE`.


```{r }
Cmcmc$run(10000, reset = FALSE, resetWAIC = FALSE)
samples_continued <- as.matrix(Cmcmc$mvSamples)
```

Let's look at the samples from the first run, and from the second.


```{r }
samples_all <- rbind(samples, samples_continued)
dim(samples_all)
```

```{r eval = FALSE}
samples_all[9995:10005,]
```

```
##        beta[1]   beta[2]       mu      sigma
##  [1,] 1.394315 0.2578996 2.709307 0.03237251
##  [2,] 1.399039 0.2502140 2.724426 0.05516978
##  [3,] 1.400995 0.2494896 2.728632 0.03437194
##  [4,] 1.400995 0.2494896 2.719170 0.03925602
##  [5,] 1.381995 0.2288096 2.741263 0.05120657
##  [6,] 1.381995 0.2780219 2.716613 0.05292085
##  [7,] 1.381995 0.2220797 2.700665 0.04997873
##  [8,] 1.381995 0.2433144 2.690273 0.08012158
##  [9,] 1.381995 0.2410102 2.669353 0.07388723
## [10,] 1.381995 0.2388559 2.696224 0.04560240
## [11,] 1.331974 0.2388559 2.694611 0.03243129
```

If using WAIC, we can look at the WAIC value from the combined runs.

```{r}
Cmcmc$getWAIC()
```

\   


## Verify this is the same as one long run

We'll double-check these samples are precisely the same, as though
we'd done one single long (un-interupted) run of the MCMC.

The code below is entirely copied from code above.


```{r message = FALSE}
###########################
### Start new R session ###
###########################

library(nimble)

code <- nimbleCode({
    mu ~ dnorm(0, 0.0001)
    sigma ~ dunif(0, 1000)
    for(i in 1:2) {
        beta[i] ~ dnorm(0, 0.0001)
    }
    for(i in 1:N) {
        epsilon[i] ~ dnorm(mu, sd = sigma)
        log(lambda[i]) <- beta[1] + beta[2] * x[i] + epsilon[i]
        y[i] ~ dpois(lambda[i])
    }
})

set.seed(0)
N <- 5
x <- rnorm(N, 0, 2)
lambda <- exp(4 + .3*x)
y <- rpois(N, lambda)

constants <- list(N = N, x = x)
data <- list(y = y)
inits <- list(mu = 0, sigma = 1, beta = rep(0,2), epsilon = rep(0, N))

Rmodel <- nimbleModel(code, constants, data, inits)

Rmodel$calculate()

conf <- configureMCMC(Rmodel, enableWAIC = useWAIC)
conf$removeSamplers("sigma")
conf$addSampler("sigma", "slice")
conf$printSamplers()

Rmcmc <- buildMCMC(conf)

Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)

set.seed(0)
Cmcmc$run(20000)
samples_all <- as.matrix(Cmcmc$mvSamples)
dim(samples_all)
samples_all[9995:10005,]

if(useWAIC)
    Cmcmc$getWAIC()
```

\   

The results look good.

Let me know if anything here is unclear.

Good luck!

Daniel


\   

\   

\   

\   

\   

\   

\   

\   

\   

\   

\   

\   













