---
title: Testing Derivatives in NIMBLE
author: Daniel Turek
output: html_document
---

<!---
danielturek.github.io/public/derivs_testing/derivs_testing..html
-->

Basic testing of `nimDerivs`.  At least, how I thought it should work.  I'm not getting the results I expected.

```{r, message = FALSE}
library(nimble)
nimbleOptions(experimentalEnableDerivs = TRUE)
```

Setup a model with some interesting posterior.  I did this to be a bi-modal posterior for `a`, idea being to get some interesting derivatives.

```{r }
code <- nimbleCode({
    a ~ dnorm(0, 1)
    y[1] ~ dnorm(a, 1)
    y[2] ~ dnorm(a^2, 1)
})
constants <- list()
data <- list(y = c(2, 4))
inits <- list(a = 0)

Rmodel <- nimbleModel(code, constants, data, inits)
```


Look at a graph of the posterior of `a`:

```{r }
xs <- seq(-4, 4, by=0.01)
ys <- numeric()
for(i in seq_along(xs)) {
    Rmodel$a <- xs[i]
    ys[i] <- Rmodel$calculate()
}

plot(xs, ys, type = 'l')
```


Just check values of the posterior against the graph, as a realit check

```{r }
Rmodel$a <- 1
Rmodel$calculate()

Rmodel$a <- 2
Rmodel$calculate()

Rmodel$a <- 3
Rmodel$calculate()
```

Now, going to make some graphs of the **derivate** of the posterior of `a`.  First using a standard R function, and calling `nimDerivs`.

```{r }
f <- function(x) {
    Rmodel$a <- x
    dOut <- nimDerivs(Rmodel$calculate(c('a','y[1]','y[2]')), order=1, wrt='a')
    d <- dOut$gradient[1,1]
    return(d)
}

ds <- numeric()
for(i in seq_along(xs)) {
    x <- xs[i]
    ds[i] <- f(x)
}
```

Plot of the posterior of `a` in black, and the derivate in blue.  Not what I was expecting.

```{r }
plot(xs, ys, type = 'l', ylim = range(c(ys,ds)))
lines(xs, ds, col = 'blue')
```

Now, try the same thing except using `derivs` inside a nimbleFunction.  Look at the same graph.

```{r }
nfDef <- nimbleFunction(
    setup = function(model, node) {},
    run = function(x = double()) {
        model[[node]] <<- x
        derivsOutput <- derivs(model$calculate(c('a','y[1]','y[2]')), order = 1, wrt = node)
        d <- derivsOutput$gradient[1,1]
        returnType(double(0))
        return(d)
    }
)

Rnf <- nfDef(Rmodel, 'a')

Rnf$run(-2)
Rnf$run(-1)
Rnf$run(0)
Rnf$run(1)
Rnf$run(2)

ds <- numeric()
for(i in seq_along(xs)) {
    x <- xs[i]
    ##ds[i] <- f(x)
    ds[i] <- Rnf$run(x)
}

plot(xs, ys, type = 'l', ylim = range(c(ys,ds)))
lines(xs, ds, col = 'blue')
```

Again, not what I was expecting (although consistent with before).  What am I doing wrong??


