---
title: "Spatial Capture-Recapture: Wolverine Example"
author: "Daniel Turek and Cyril Milleret"
date: 17 October 2019
output: html_document
---

<!---
http://danielturek.github.io/public/scr/wolverine_example 
-->



<!--
Back to [NIMBLE Vignettes](https://danielturek.github.io/public/nimble_vignettes.html) 
-->


\  

This example demonstrates the Wolverine example model, as in
"Efficient MCMC for Spatial Capture-Recapture Models" (Turek *et al*,
*submitted*).

Specifically, we implement the final version of the
model and MCMC.  Details of the functions and procedure are provided therein.



\   

### Load Libraries 

```{r message = FALSE}
library(nimble)
```



\  

### Define Custom Distributions

We define two custom distributions for use in the model.

`dBernoulliVector3` implements the vectorized likelihood calculations, only
for individuals in the population, and for a restricted set of traps.

`dHabitat` enforces the habitat mask to restrict activity center
locations to the habitable regions of the study area.

We also register these distributions for use in `nimble` models, using
`registerDistributions`.  But even without doing this, `nimble` would
find them at the time of model building.


```{r eval = FALSE}
dBernoulliVector3 <- nimbleFunction(
    run = function(
        x = double(1), pZero = double(0), sxy = double(1), sigma = double(0),
        nbDetections = double(0), yDets = double(1), detector.xy = double(2),
        trials = double(1), detectorIndex = double(2), nDetectors = double(1),
        ResizeFactor = double(0, default = 1), maxNBDets = double(0),
        habitatID = double(2), indicator = double(0), log = integer(0)) {
        ##
        numTrials <- length(trials)
        ##
        if(indicator == 0) {
            if(nbDetections == 0) {
                if(log == 0) return(1)
                else return(0)
            } else {
                if(log == 0) return(0)
                else return(-Inf)
            }
        }
        ##
        sxyID <- habitatID[trunc(sxy[2]/ResizeFactor)+1, trunc(sxy[1]/ResizeFactor)+1]
        index <- detectorIndex[sxyID, 1:nDetectors[sxyID]]
        ##
        y <- nimNumeric(length = numTrials, value = 0, init = TRUE)
        ##
        if(nbDetections > 0) {
            for(r in 1:nbDetections) {
                y[yDets[r]] <- x[r] 
                if(sum(yDets[r]==index) == 0) {
                    if(log == 0) return(0)
                    else return(-Inf)
                } 
            } 
        }
        ##
        alpha <- -1/(2*sigma^2)
        lp <- 0
        count <- 1 
        index1 <- c(index, 0)   # necessary
        ##
        for(r in 1:numTrials) {
            if(index1[count] == r) {
                d2 <- (detector.xy[r,1] - sxy[1])^2 + (detector.xy[r,2] - sxy[2])^2
                p <- pZero * exp(alpha * d2)
                lp <- lp + dbinom(y[r], prob = p, size = trials[r], log = TRUE)
                count <- count + 1
            }
        }
        ##
        returnType(double(0))
        return(lp)
    }
)


dHabitat <- nimbleFunction(
    run = function(x = double(0), sxy = double(1), lower = double(1),
        upper = double(1), habitat = double(2), log = double()) {
        if(sxy[1] < lower[1]) return(-Inf)
        if(sxy[1] > upper[1]) return(-Inf)
        if(sxy[2] < lower[2]) return(-Inf)
        if(sxy[2] > upper[2]) return(-Inf)
        ## habitat constraint:
        if(habitat[trunc(sxy[2])+1, trunc(sxy[1])+1] == 0) return(-Inf)
        returnType(double())
        return(0)
    }
)


registerDistributions(
    list(
        dBernoulliVector3 = list(
            BUGSdist = "dBernoulliVector3(pZero, sxy, sigma, nbDetections, yDets, 
                detector.xy, trials, detectorIndex, nDetectors,
                ResizeFactor, maxNBDets, habitatID, indicator)",
            types = c("value = double(1)", "pZero = double(0)","sxy = double(1)",
                "sigma = double(0)", "nbDetections = double(0)",
                "yDets = double(1)", "detector.xy = double(2)",
                "trials = double(1)","detectorIndex = double(2)" ,
                "nDetectors = double(1)", "ResizeFactor = double(0)",
                "maxNBDets = double(0)","habitatID= double(2)",
                "indicator = double(0)"),
            mixedSizes = TRUE)))


registerDistributions(
    list(
        dHabitat = list(
            BUGSdist = "dHabitat(sxy, lower, upper, habitat)",
            types = c("value = double(0)", "sxy = double(1)", "lower = double(1)",
                "upper = double(1)", "habitat = double(2)"),
            mixedSizes = TRUE)
    )
)
```

\   

### Define Model Structure

Next we define the model struture.

This uses the two custom distributions defined above.


```{r }
code <- nimbleCode({
    ## priors
    psi ~ dunif(0, 1)
    sigma ~ dunif(0, 50)
    p0 ~ dunif(0, 1)
    ##
    for(i in 1:n.individuals) {
        ## initial ACs
        sxy[i,1] ~ dunif(0, x.max)
        sxy[i,2] ~ dunif(0, y.max)
        ## habitat constraint
        ones[i] ~ dHabitat(sxy = sxy[i,1:2],
                           lower = lowerCoords[1:2], 
                           upper = upperCoords[1:2],
                           habitat = habitat.mx[1:y.max,1:x.max])
        ## latent dead/alive indicators
        z[i] ~ dbern(psi)
        ## likelihood
        y[i, 1:nMaxDetectors] ~ dBernoulliVector3(
            pZero = p0, sxy = sxy[i,1:2], sigma = sigma,
            nbDetections[i], yDets = yDets[i,1:nMaxDetectors],
            detector.xy =  detector.xy[1:n.detectors,1:2],
            trials = trials[1:n.detectors],
            detectorIndex = detectorIndex[1:n.cells,1:maxNBDets], 
            nDetectors = nDetectors[1:n.cells],  ResizeFactor = ResizeFactor,
            maxNBDets = maxNBDets, habitatID = habitatIDDet[1:y.maxDet,1:x.maxDet],
            indicator = z[i])
    }
    ## derived quantity: total population size
    N <- sum(z[1:n.individuals])
})
```

\   

### Load and Process Data

We load the Wolverine example data.

The data file can be [downloaded here](http://danielturek.github.io/public/scr/wolverine_data.RData).

We also create the `code`, `constants`, `data`, and `inits` for later use in 
the function `nimbleModel`. 

```{r }
load("wolverine_data.RData")

data <- list(y = wolverine_input$y,
             z = wolverine_input$z,
             detector.xy = wolverine_input$detector.xy,
             habitat.mx = wolverine_input$habitat.mx,
             ones = wolverine_input$OK,
             lowerCoords = c(0,0),
             upperCoords = c(
                 dim(wolverine_input$habitat.mx)[2],
                 dim(wolverine_input$habitat.mx)[1]),
             trials = rep(1, dim(wolverine_input$detector.xy)[1]))

constants <- list(n.individuals = wolverine_input$n.individuals,
                  n.detectors = dim(wolverine_input$detector.xy)[1],
                  y.max = dim(wolverine_input$habitat.mx)[1],
                  x.max = dim(wolverine_input$habitat.mx)[2])

inits <- list(sxy = wolverine_inits$sxy,
              z = wolverine_inits$z,
              p0 = 0.05,
              psi = 0.5,
              sigma = 6)
```

\   

Now we do some processing of the data, for use in this version
of the model.

The function `GetDetectorIndex` identifies the set of detectors that
are within a certain distance $d_{max}$ from each habitat cell
center. The value of $d_{max}$ should be large enough so that for any
particular individual, the `sxy` values of the initial activity center
restrict the calculation of $p_0$ to all detectors with positive detections. Here, we use $d_{max}=38$.

Because of the large dimensions of the habitat matrix, we resize it to
lower dimensions. This reduces the number of habitat cells for which
we have to identify the set of detectors that are $d_{max}$ from the
cell center. We resize the habitat matrix by a factor of 24, which
corresponds to the the `fact` argument used internally in `raster::disaggregate`. The goal is to create the object `DetectorIndex$detectorIndex` of the smallest dimension possible. 

The `GetDetectorIndex` function is defined in the auxiliary script
`wolverine_functions.R`, which is [available for download here](http://danielturek.github.io/public/scr/wolverine_functions.R).

```{r, message = FALSE}
source("wolverine_functions.R")
library(raster)
```


```{r, fig.width = 6, fig.height = 7}
set.seed(5)
DetectorIndex <- GetDetectorIndex(habitat.mx = data$habitat.mx,
                                  detectors.xy = data$detector.xy,
                                  maxDist = 38,
                                  ResizeFactor = 24)

constants$y.maxDet <- dim(DetectorIndex$habitatID)[1]
constants$x.maxDet <- dim(DetectorIndex$habitatID)[2]
constants$ResizeFactor <- DetectorIndex$ResizeFactor
constants$n.cells <- dim(DetectorIndex$detectorIndex)[1]
constants$maxNBDets <- DetectorIndex$maxNBDets

data$detectorIndex <- DetectorIndex$detectorIndex
data$nDetectors <- DetectorIndex$nDetectors
data$habitatIDDet <- DetectorIndex$habitatID
```

\  

Last, we re-express the detection matrix `y` as `ySparseDets`, to reduce its size. We use a new representation, where each row (corresponding to one individual) contains the detector identification numbers (values of *r*) which detected that individual. 

A second matrix of identical dimension was also created, containing the number of detections occurring at each detector. The second matrix would be necessary for modelling non-binary detections. 


```{r }
y <- wolverine_input$y
nbDetections <- apply(y,1, function(x) length(which(x>0)))

ySparseDets <- matrix(-1, nrow=dim(y)[1], ncol=max(nbDetections))

for(i in 1:dim(y)[1]) {
    if(nbDetections[i]>0) {
        ySparseDets[i, 1:nbDetections[i]] <- which(y[i,]>0)
    }
}

ySparse <- matrix(-1, nrow=dim(y)[1], ncol=max(nbDetections))

for(i in 1:dim(y)[1]) {
    if(nbDetections[i] > 0) {
        ySparse[i, 1:nbDetections[i]] <- y[i,which(y[i,] > 0)]
    }
}

data$y <- ySparse
data$yDets <- ySparseDets
data$nbDetections <- nbDetections

constants$nMaxDetectors <- max(nbDetections)
```

\  


### Create NIMBLE Model

Now we create the `nimble` model object, using the model structure
defined in `code`, and the constants, data, and initial values.


```{r eval = FALSE}
Rmodel <- nimbleModel(code, constants, data, inits)
```

\   

### Configure and Build MCMC

We configure an MCMC algorithm to the `Rmodel` model object.

We assign MCMC monitors to $N$, $\sigma$, and $p_0$.

We also remove the univariate Metropolis-Hastings samplers which were
assigned (by default) to each AC component, the `sxy` variables.
Rather, we add joint Metropolis-Hastings samplers (`RW_block`)
samplers to each $x$ and $y$ coordinate pair `sxy[i, 1:2]`.


```{r eval = FALSE}
conf <- configureMCMC(Rmodel, monitors = c("N", "sigma", "p0"))

conf$removeSamplers("sxy")

ACnodes <- paste0("sxy[", 1:constants$n.individuals, ", 1:2]")

for(node in ACnodes) {
    conf$addSampler(target = node,
                    type = "RW_block",
                    control = list(adaptScaleOnly = TRUE),
                    silent = TRUE)
}

Rmcmc <- buildMCMC(conf)
```

\   


### Compile and Run MCMC

Compile both the model and MCMC objects.

Execute the compiled MCMC for 10,000 iterations.

```{r eval = FALSE}
Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)

set.seed(0)
time <- system.time(
    samples <- runMCMC(Cmcmc, niter = 10000)
)
```


```{r eval = FALSE, echo = FALSE}
save(samples, time, file = "wolverine_samples.RData")
```


```{r echo = FALSE}
load("wolverine_samples.RData")
```

\   

### Results

MCMC runtime, in minutes:


```{r }
time[3] / 60
```


Posterior effective sample size (ESS) resulting from our
10,000 posterior samples

```{r }
library(coda)
ESS <- coda::effectiveSize(samples)
ESS
```

MCMC efficiency for each parameter: the rate of generating effectively
independent posterior samples, per second of MCMC runtime:

```{r }
ESS/time[3]
```

Summary of posterior distributions for each parameter:

```{r }
round(samplesSummary(samples), 2)
```

Examine traceplots and posterior distributions:

```{r }
library(basicMCMCplots)
basicMCMCplots::chainsPlot(samples)
```


\  

\  

\  

\  

\  

\  

\  

\  

\  

\  

\  

\  

\  

\  

\  

