---
title: Toggle MCMC Samplers
author: Daniel Turek
output: html_document
---

<!---
danielturek.github.io/public/folder/fileName.html
-->

This will demonstrate how to toggle MCMC samplers on/off.  It's a fairly custom approach, not really intended for mainstream use, but the techniques could be refined.  In any case, it should get you going.
 
I would be happy to be kept in the loop, regarding this research.

## Preliminaries

Load NIMBLE library.

```{r, message = FALSE}
library(nimble)
```

You'll need this sampler definition.

It will generate a warning when you define it.  You can thank Chris for that  ;)

```{r message = FALSE}
sampler_toggle <- nimbleFunction(
    contains = sampler_BASE,
    setup = function(model, mvSaved, target, control) {
        type <- control$type
        nested_sampler_name <- paste0('sampler_', type)
        control_new <- nimbleOptions('MCMCcontrolDefaultList')
        control_new[[names(control)]] <- control
        nested_sampler_list <- nimbleFunctionList(sampler_BASE)
        nested_sampler_list[[1]] <- do.call(nested_sampler_name, list(model, mvSaved, target, control_new))
        toggle <- 1
    },
    run = function() {
        if(toggle == 1)
            nested_sampler_list[[1]]$run()
    },
    methods = list(
        reset = function()
            nested_sampler_list[[1]]$reset()
    )
)
```

## Setting up the model and MCMC

This next section only needs to be done once.  There are several important specifics here, and generally all steps must be done, in order.

The model is a simplification of yours, but it should work fine adding additional layers above `y.censored` and `X`.

```{r message=FALSE, warning=FALSE}
code <- nimbleCode({ 
    ##q[1:N,1:N]  ~ dwish(R = aq[1:N,1:N], df = bq)
    ##Q[1:N,1:N] <- inverse(q[1:N,1:N])
    ##X.mod[1:N] ~ dmnorm(muf[1:N],prec = pf[1:N,1:N])
    ## add process error
    X[1:N]  ~ dmnorm(X.mod[1:N],prec = q[1:N,1:N])
    ## Analysis
    y.censored[1:N] ~ dmnorm(X[1:N], prec = r[1:N,1:N]) 
    for(i in 1:N){
        y.ind[i] ~ dconstraint(y.censored[i] > 0)
    }
})

N <- 5   ## works with other values of N

constants <- list(N=N, X.mod=rep(10,N), q=diag(N), r=diag(N))

## the values of y.ind and y.censored are just placeholders...
## BUT, important, y.censored *must* be specified as data at this point.
## it's a long story why, has to do with the initializeModel routine
## at the beginning of MCMC execution
data <- list(y.ind=rep(1,N), y.censored=rep(10,N))  

inits <- list(X=rep(10,N))

Rmodel <- nimbleModel(code, constants, data, inits)

conf <- configureMCMC(Rmodel, print=TRUE)

## important!
## this is needed for correct indexing later
samplerNumberOffset <- length(conf$getSamplers())

for(i in 1:N) {
    node <- paste0('y.censored[',i,']')
    conf$addSampler(node, 'toggle', control=list(type='RW'))
    ## could instead use slice samplers, or any combination thereof, e.g.:
    ##conf$addSampler(node, 'toggle', control=list(type='slice'))
}

conf$printSamplers()

conf$printMonitors()

## can monitor y.censored, if you wish, to verify correct behaviour
conf$addMonitors('y.censored')

Rmcmc <- buildMCMC(conf)

Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)
```

## Repeated steps with new datasets

The steps hereafter can be repeated with new datasets (as
demonstrated)

\   

#### New Dataset 1 

Mix of data and non-data.

```{r }
y.ind <- c(1, 1, 1, 0, 0)

## important!
## change: rather than providing NA for the non-data values (those to be sampled),
## you'll have to provide some values here.
## that's because we effectively disabled the model initialization routine earlier
y.censored <- c(9, 9, 11, 20, 20)

Cmodel$y.ind <- y.ind
Cmodel$y.censored <- y.censored

for(i in 1:N) {
    ## ironically, here we have to "toggle" the value of y.ind[i]
    ## this specifies that when y.ind[i] = 1,
    ## indicator variable is set to 0, which specifies *not* to sample
    valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[samplerNumberOffset+i]], 'toggle', 1-y.ind[i])
}

niter <- 1000
set.seed(0)
samples <- runMCMC(Cmcmc, niter, progressBar=FALSE)

head(samples)
tail(samples)
```

\   

#### New Dataset 2 

Everything is data.

```{r }
y.ind <- c(1, 1, 1, 1, 1)   ## everything is data

y.censored <- c(9, 9, 11, 11, 12)
 
Cmodel$y.ind <- y.ind 
Cmodel$y.censored <- y.censored
 
for(i in 1:N) {
    valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[samplerNumberOffset+i]], 'toggle', 1-y.ind[i])
}

niter <- 1000 
set.seed(0)
samples <- runMCMC(Cmcmc, niter, progressBar=FALSE)

head(samples)
tail(samples)
```

\   

#### New Dataset 3

Nothing is data.

```{r }
y.ind <- c(0, 0, 0, 0, 0)   ## nothing is data

## again, you'll have to provide actual numeric (initial) values
## for the nodes that will be sampled:
y.censored <- c(20, 20, 20, 20, 20)

Cmodel$y.ind <- y.ind
Cmodel$y.censored <- y.censored

for(i in 1:N) {
    valueInCompiledNimbleFunction(Cmcmc$samplerFunctions[[samplerNumberOffset+i]], 'toggle', 1-y.ind[i])
}

niter <- 1000
set.seed(0)
samples <- runMCMC(Cmcmc, niter, progressBar=FALSE)

head(samples)
tail(samples)
```

Send along any questions or comments.

-Daniel

\  

\  

\  

\  





