
---
title: Example of Sampler Timings
author: Daniel Turek
date: 11 July 2016
output: html_document
---

```{r, message = FALSE}
library(nimble)
```

Define a model with one node.

```{r, message = FALSE }
code <- nimbleCode({
    a ~ dnorm(0, 1)
})
constants <- list()
data <- list()
inits <- list(a = 0)

Rmodel <- nimbleModel(code, constants, data, inits)
```

We'll add three samplers to node `a`.

The first RW sampler is not adaptive, so it should be fast.

The second RW sampler is adaptive with adaptInterval = 10, very frequent, so it should be slower.

The third is a slice sampler, which should be the slowest.

```{r }
conf <- configureMCMC(Rmodel, nodes = NULL)

conf$addSampler(target='a', type='RW', control = list(adaptive=FALSE), print=FALSE)
conf$addSampler(target='a', type='RW', control = list(adaptive=TRUE, adaptInterval=10), print=FALSE)
conf$addSampler(target='a', type='slice', print=FALSE)

conf$printSamplers()
```

Build, compile, and run MCMC.

When you run the MCMC using `mcmc$run(...)`, add the extra argument `time = TRUE`, which will make it record sampler timings.


```{r }
Rmcmc <- buildMCMC(conf)
Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)
system.time(Cmcmc$run(1000000, time = TRUE))
```

Now we'll extract the times.  It's a vector, with length the number of samplers (in this case, three).  Each number corresponds to the number of seconds (total) spent in each sampler.


```{r }
times <- Cmcmc$getTimes()
times
```

