---
title: Saving Model and MCMC State
author: Daniel Turek
output: html_document
---


<!---
danielturek.github.io/public/saveMCMCstate/saveMCMCstate
--->


Back to [NIMBLE Vignettes](https://danielturek.github.io/public/nimble_vignettes.html)

\    

This demonstrates how to "save" the model state, and internal state of
an MCMC algorithm to disk.

Then, R can be restarted, and this saved "state" can be reloaded into a new (structurally identical) model and MCMC algorithm, allowing the original sampling chain to be "resumed" where it left off.

\   

### Function Definitions

\   

```{r }
getStateVariableNames <- function(samplerDef) {
    resetMethod <- body(samplerDef$reset)
    stateVars <- character()
    if(resetMethod[[1]] != '{') stop('something wrong')
    numLines <- length(resetMethod)
    for(i in 1:numLines) {
        if(i == 1) next
        thisLine <- resetMethod[[i]]
        if(thisLine[[1]] == '<<-') {
            LHS <- thisLine[[2]]
            if(!is.name(LHS)) stop('haven\'t dealt with non-name-LHS case yet')
            stateVars <- c(stateVars, as.character(LHS))
        }
        if('my_calcAdaptationFactor' %in% all.names(thisLine)) {
            stateVars <- c(stateVars, 'my_calcAdaptationFactor')
        }
    }
    setupMethod <- body(samplerDef$setup)
    if('empirSamp' %in% all.names(setupMethod)) stateVars <- c(stateVars, 'empirSamp')
    return(stateVars)
}

getModelState <- function(model) {
    modelVarNames <- model$getVarNames()
    modelVarValuesList <- vector('list', length(modelVarNames))
    names(modelVarValuesList) <- modelVarNames
    for(var in modelVarNames) {
        modelVarValuesList[[var]] <- model[[var]]
    }
    return(modelVarValuesList)
}

getMCMCstate <- function(conf, mcmc) {
    stateVarNamesList <- vector('list', length(conf$samplerConfs))
    mcmcStateValuesList <- vector('list', length(conf$samplerConfs))
    for(i in seq_along(conf$samplerConfs)) {
        samplerDef <- conf$getSamplerDefinition(i)
        theseStateNames <- getStateVariableNames(samplerDef)
        theseStateValuesList <- vector('list', length(theseStateNames))
        names(theseStateValuesList) <- theseStateNames
        for(j in seq_along(theseStateNames)) {
            if(is.nf(mcmc)) {
                if(theseStateNames[j] == 'my_calcAdaptationFactor') {
                    theseStateValuesList[[j]] <- list(timesAdapted = mcmc$samplerFunctions$contentsList[[i]][[theseStateNames[j]]]$timesAdapted,
                                                      gamma1 = mcmc$samplerFunctions$contentsList[[i]][[theseStateNames[j]]]$gamma1)
                } else
                    theseStateValuesList[[j]] <- mcmc$samplerFunctions$contentsList[[i]][[theseStateNames[j]]]
            }
            if(is.Cnf(mcmc)) {
                if(theseStateNames[j] == 'my_calcAdaptationFactor') {
                    theseStateValuesList[[j]] <- list(timesAdapted = valueInCompiledNimbleFunction(mcmc$Robject$samplerFunctions$contentsList[[i]]$my_calcAdaptationFactor$.CobjectInterface, 'timesAdapted'),
                                                      gamma1 = valueInCompiledNimbleFunction(mcmc$Robject$samplerFunctions$contentsList[[i]]$my_calcAdaptationFactor$.CobjectInterface, 'gamma1'))
                } else
                    theseStateValuesList[[j]] <- valueInCompiledNimbleFunction(mcmc$samplerFunctions[[i]], theseStateNames[j])
            }
        }
        mcmcStateValuesList[[i]] <- theseStateValuesList
    }
    return(mcmcStateValuesList)
}

getWAICstate <- function(mcmc) {
    waicStateNames <- c('delta1pWAICmat', 'delta2pWAICmat', 'finalized', 'logProbMat', 'lppdCurSumMat', 'lppdSumMaxMat',
       'mcmcIter', 'meanpWAICmat', 'sspWAICmat')
    waicStateList <- vector('list', length(waicStateNames))
    names(waicStateList) <- waicStateNames
    for(nm in waicStateNames) {
        if(is.nf(mcmc)) {
            waicStateList[[nm]] <- mcmc$waicFun[[1]][[nm]]
        }
        if(is.Cnf(mcmc)) {
            waicStateList[[nm]] <- valueInCompiledNimbleFunction(mcmc$waicFun[[1]], nm)
        }
    }
    return(waicStateList)
}

setModelState <- function(model, modelState) {
    modelVarNames <- model$getVarNames()
    if(!identical(sort(modelVarNames), sort(names(modelState)))) stop('saved model variables don\'t agree')
    for(var in modelVarNames) {
        model[[var]] <- modelState[[var]]
    }
    invisible(model$calculate())
}

setMCMCstate <- function(conf, mcmc, mcmcState) {
    if(length(mcmcState) != length(conf$samplerConfs)) stop('saved mcmc samplers don\'t agree')
    for(i in seq_along(conf$samplerConfs)) {
        theseStateValuesList <- mcmcState[[i]]
        for(j in seq_along(theseStateValuesList)) {
            samplerStateName <- names(theseStateValuesList)[j]
            if(is.nf(mcmc)) {
                if(samplerStateName == 'my_calcAdaptationFactor') {
                    mcmc$samplerFunctions$contentsList[[i]]$my_calcAdaptationFactor$timesAdapted <- theseStateValuesList[[j]]$timesAdapted
                    mcmc$samplerFunctions$contentsList[[i]]$my_calcAdaptationFactor$gamma1 <- theseStateValuesList[[j]]$gamma1
                } else {
                    mcmc$samplerFunctions$contentsList[[i]][[samplerStateName]] <- theseStateValuesList[[samplerStateName]]
                }
            }
            if(is.Cnf(mcmc)) {
                if(samplerStateName == 'my_calcAdaptationFactor') {
                    invisible(valueInCompiledNimbleFunction(mcmc$Robject$samplerFunctions$contentsList[[i]]$my_calcAdaptationFactor$.CobjectInterface, 'timesAdapted', theseStateValuesList[[j]]$timesAdapted))
                    invisible(valueInCompiledNimbleFunction(mcmc$Robject$samplerFunctions$contentsList[[i]]$my_calcAdaptationFactor$.CobjectInterface, 'gamma1', theseStateValuesList[[j]]$gamma1))
                } else {
                    invisible(valueInCompiledNimbleFunction(mcmc$samplerFunctions[[i]], samplerStateName, theseStateValuesList[[samplerStateName]]))
                }
            }
        }
    }
}

setWAICstate <- function(mcmc, waicState) {
    for(nm in names(waicState)) {
        if(is.nf(mcmc)) {
            mcmc$waicFun[[1]][[nm]] <- waicState[[nm]]
        }                       
        if(is.Cnf(mcmc)) {
            invisible(valueInCompiledNimbleFunction(mcmc$waicFun[[1]], nm, waicState[[nm]]))
        }
    }
}
```


\   

### Usage Example

\   

```{r, message = FALSE}
library(nimble)
```

\  
 
Define a NIMBLE model object, and an MCMC algorithm to operate on it.
Optionally, choose to calculate WAIC.

The saving / loading state will work for *most any* NIMBLE MCMC algorithms, but
if you have questions about your use case, then
please email Daniel at `dbt1@williams.edu`.

\  

```{r }
useWAIC <- TRUE

code <- nimbleCode({
    for(i in 1:N) {
        x[i] ~ dnorm(0, 1)
    }
    b ~ dbern(0.5)
    c ~ dnorm(0, 1)
    y ~ dnorm(c, 1)
    p ~ dnorm(0, 1)
})

N <- 8
constants <- list(N = N)
data <- list(y = 0)
inits <- list(x = rep(0, N),
              b = 0,
              c = 0,
              p = 0)

## create model object
Rmodel <- nimbleModel(code, constants, data, inits)
Rmodel$calculate()

## configure & build MCMC
conf <- configureMCMC(Rmodel, enableWAIC = useWAIC)
conf$removeSamplers(c('b', 'x'))
conf$addSampler('b', 'binary')
conf$addSampler('x[1]', 'RW')
conf$addSampler('x[2]', 'slice')
conf$addSampler('x[3:5]', 'RW_block')
conf$addSampler('x[6:8]', 'AF_slice')
conf$printSamplers()
Rmcmc <- buildMCMC(conf)

## compile
Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)

## execute MCMC for 10000 iterations
set.seed(0)
Cmcmc$run(10000)
```

\  
 
Now, we'll "extract" and save the model state, and the internal state
of the MCMC.

For saving the MCMC state, you'll need both the compiled MCMC object
(`Cmcmc`), and also the MCMC configuration object (`conf`).

\  

```{r }
stateList <- list(modelState = getModelState(Cmodel),
                  mcmcState = getMCMCstate(conf, Cmcmc))
if(useWAIC)
    stateList[['waicState']] <- getWAICstate(Cmcmc)

saveRDS(stateList, file = '~/temp/state.rds')
```

\  
 
Here, you could restart R.

Now we'll make new (compiled) model and MCMC objects, then load this
previous "state" back into them.

\  

```{r }
## construct an *identical* model object
Rmodel2 <- nimbleModel(code, constants, data, inits)

## construct an *identical* MCMC algorithm
conf2 <- configureMCMC(Rmodel2, enableWAIC = useWAIC)
conf2$removeSamplers(c('b', 'x'))
conf2$addSampler('b', 'binary')
conf2$addSampler('x[1]', 'RW')
conf2$addSampler('x[2]', 'slice')
conf2$addSampler('x[3:5]', 'RW_block')
conf2$addSampler('x[6:8]', 'AF_slice')
conf2$printSamplers()
Rmcmc2 <- buildMCMC(conf2)

## compile
Cmodel2 <- compileNimble(Rmodel2)
Cmcmc2 <- compileNimble(Rmcmc2, project = Rmodel2)
```

\  
 
Now, we'll "restore" the previous state into the new model (`Cmodel2`)
and the new MCMC (`Cmcmc2`).

\  

```{r }
## load the saved "state" file
stateList <- readRDS(file = '~/temp/state.rds')

modelState <- stateList$modelState
mcmcState <- stateList$mcmcState
if(useWAIC)
    waicState <- stateList$waicState

## restore the saved "state" into the new model and new MCMC
setModelState(Cmodel2, modelState)
setMCMCstate(conf2, Cmcmc2, mcmcState)
if(useWAIC)
    setWAICstate(Cmcmc2, waicState)
```

\  
 
Finally, now you can "resume" the MCMC run, exactly where it left off.

**Important:** Make certain to use the argument `reset = FALSE`, so the
  MCMC state is *not reset* to its initial settings for a new, fresh,
  MCMC run. If calculating WAIC, also set `resetWAIC = FALSE`.

This option, `reset = FALSE`, is only available through the
`mcmc$run(...)` invocation of the MCMC algorithm.

\  

```{r }
## continue MCMC run
Cmcmc2$run(10000, reset = FALSE, resetWAIC = FALSE)
```

\   

That's it!  You did it!



\   

\   

\   

\   

\   

\   

\   

\   

\   

\   

\   

\   

\   

\   

\   

\   

\   

\   



