---
title: R Outcalls to Sparse Matrix Operations
author: Daniel Turek
date: 11 January 2019
output: html_document
---

<!---
danielturek.github.io/public/sparseMatrixCall/sparseMatrixCall.html
-->

Back to [NIMBLE Vignettes](https://danielturek.github.io/public/nimble_vignettes.html)

\    


Here we'll show how to call the `Matrix::tcrossprod` function from the
`Matrix` package inside a compiled `nimbleFunction`.

\   

### Setup

A function to build a full (dense) matrix, from the component `i` (row
indices), `j` (column indices), and `x` (non-zero values) sparse representation.

```{r }
makeMatrix <- function(i, j, x, nr, nc, ijx) {
    ## ijx is an (optional, alternate) representation of {i,j,x} as a 3-column array
    if(!missing(ijx))   { i <- ijx[,1]; j <- ijx[,2]; x <- ijx[,3] }
    if(length(i) != length(j) | length(i) != length(x)) stop()
    if(missing(nr)) nr <- max(i)
    if(missing(nc)) nc <- max(j)
    mat <- array(0, c(nr, nc))
    for(ind in seq_along(i))   mat[i[ind], j[ind]] <- x[ind]
    return(mat)
}
```

\   

Try using `makeMatrix`, and also the standard `tcrossprod` function from `base`.

```{r }
i <- c(1,3,7,7)
j <- c(1,1,3,4)
x <- c(1,2,3,4)

A <- makeMatrix(i, j, x)
A

A %*% t(A)    ## what tcrossprod(A) does
base::tcrossprod(A)    ## standard tcrossprod from base package
```
\   

### Matrix Package

Now we'll do the same `tcrossprod` calculation using the sparse
representation and the `Matrix::tcrossprod` function.

```{r }
library(Matrix)

Asparse <- Matrix::sparseMatrix(i, j, x = x)
Asparse

Matrix::tcrossprod(Asparse)
```

\   

### Calling `Matrix::tcrossprod` from NIMBLE

Now, we're ready to build a wrapper function for the
`Matrix::tcrossprod` function, then use `nimbleRcall` to register
this as a `nimbleFunction`, that can be used inside other compiled `nimbleFunctions`.


```{r }
R_sparse_crossprod <- function(i, j, x) {
    require(Matrix)
    Asparse <- Matrix::sparseMatrix(i, j, x = x)
    ans.dsCMatrix <- Matrix::tcrossprod(Asparse)
    ans.dgTMatrix <- as(ans.dsCMatrix, 'dgTMatrix')
    i <- ans.dgTMatrix@i + 1
    j <- ans.dgTMatrix@j + 1
    x <- ans.dgTMatrix@x
    ijx <- cbind(i, j, x)
    return(ijx)
}
```

Check regular R execution of our R function:

```{r }
ijx <- R_sparse_crossprod(i, j, x)
ijx

makeMatrix(ijx = ijx)
```

\   

Now, we're ready to call this R function `R_sparse_crossprod` from
within a `nimbleFunction`.

```{r , message = FALSE}
library(nimble)
```

Now we register use of the `R_sparse_crossprod` function, essentially just like any
other `nimbleFunction`, using `nimbleRcall`.

See `help(nimbleRcall)`.

```{r }
nimble_sparse_crossprod <- nimbleRcall(
    prototype = function(i = double(1), j = double(1), x = double(1)) {},
    returnType = double(2),
    Rfun = 'R_sparse_crossprod'
)
```

For example, here we'll define another `nimbleFunction` (e.g.,
this could represent a SGV density evaluation) which will internally
use `nimble_sparse_crossprod` in order to call `Matrix::tcrossprod`.

This function will be R-executable, and also compilable and executable.

```{r }
Rnf <- nimbleFunction(
    run = function(i = double(1), j = double(1), x = double(1)) {
        ijx <- nimble_sparse_crossprod(i, j, x)
        returnType(double(2))
        return(ijx)
    }
)
     
Cnf <- compileNimble(Rnf)

Rnf(i, j, x)
Cnf(i, j, x)

makeMatrix(ijx = Rnf(i, j, x))
makeMatrix(ijx = Cnf(i, j, x))
```

\   

### Timing Comparisons

Some brief timing comparisons, between regular R execution of
`R_sparse_crosspord` wrapper for `Matrix::tcrossprod`, compared
against call-back from compiled execution though NIMBLE using the
`Cnf` compiled `nimbleFunction`.

```{r }
compare <- function(n) {
    set.seed(0)
    i <- ceiling(runif(n, 1, n))
    j <- ceiling(runif(n, 1, n))
    x <- 1:n
    R.time <- system.time(R.ans <- R_sparse_crossprod(i, j, x))[3]
    C.nimble.time <- system.time(C.nimble.ans <- Cnf(i, j, x))[3]
    if(!identical(unname(R.ans), C.nimble.ans)) stop('error')
    return(data.frame(n = n,
                      R.time = as.numeric(R.time),
                      C.nimble.time = as.numeric(C.nimble.time)))
}

p <- 6
nvals <- 10^(1:p)
as.data.frame(t(sapply(nvals, compare)))
```

\   

\   

\   

\   

\   

\   

\   

\   

\   

\   

\   

\   



